Evaluation of classical correlation functions from 2/3D images on CPU and GPU architectures: introducing CorrelationFunctions.jl

### 2023-04-26 paper-correlations.tex
Fix a typo
Rework map section

The next function is a cluster function $C_2$.
\begin{definition}
  Cluster function $C_2^{(i)}(\bm{r})$ equals to probability that both
  ends of a vector $\bm{r}$ lie in the same cluster when the vector is
  randomly thrown into the sample. A cluster is a set of points of the same
  phase $i$ where any to points can be connected by a path lying entirely in
  that phase.
\end{definition}

The next two functions describe surface of a sample. These functions are
surface-surface and surface-void functions. For homogenous media their
definitions are:
\begin{definition}
  Surface-surface function $F_{ss}^{(i)}(\bm{r})$ equals to probability that
  both ends of a vector $\bm{r}$ touch the interface of phase $i$ when the
  vector is randomly thrown into the sample.
\end{definition}
and
\begin{definition}
  Surface-void function $F_{sv}^{(i)}(\bm{r})$ equals to probability that
  one end of a vector $\bm{r}$ touches the interface of phase $i$ and the
  other end touches the void phase when the vector is randomly thrown into the
  sample.
\end{definition}
If we introduce an interface indicator function
\begin{equation*}
  M^{(i)}(\bm{r}) = | \nabla I^{(i)}(\bm{r}) |
\end{equation*}
we can define these functions mathematically as
\begin{align}
  F_{ss}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) M^{(i)}(\bm{x} +
  \bm{r}) \rangle \label{eq:fss-def} \\
  F_{sv}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) I^{(void)}(\bm{x}
  + \bm{r}) \rangle \label{eq:fsv-def}
\end{align}
As always, there are versions of these functions applicable to scalar argument
for isotropic media.

@@

Finally, a chord length function $p^{(i)}(r)$ is defined as follows:
\begin{definition}
$p^{(i)}(r)dz$ = Probability of finding a chord of length between $r$ and $r+dr$
in phase $i$. Chords are all of the line segments between intersections of an
infinitely long line with the two-phase interface.
\end{definition}

\verb+CorrelationFunctions.jl+ is capable to compute all these
functions. Usually there are two implementations for a function. The first
implementation computes a correlation function in specific predefined directions
(i.e. orientation of test vectors thrown into a sample is fixed). There are one
predefined direction in 1D case, four directions in 2D case and thirteen
directions in 3D case. This implementation is somewhat slow but requires much
less memory than the second one. The second implementation computes a
correlation function in all directions and is faster but requires much memory.
These implementations are contained in \code{Directional} and \code{Map} modules
of the package correspondingly.

@@

\section{The map approach.}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of interface indicator $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into sample). If we want to calculate $S_2$ function in periodic mode we
need only to apply \cref{eq:s2-ft} to an input to obtain the result because DFT
implicitly continues a signal to infinity \textcolor{red}{Does this need a
clarification?}. In non-periodic mode we need to pad an input with zeros to at
least twice the size minus one for each dimension. In periodic mode a total
number of trials is independent of $r$ and equals to the number of elements in
an input array, or otherwise is calculated as described
\textcolor{red}{Here}. The following algorithm summarizes all said above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}| / N$
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation if $O(M \log M)$ where $M$ is the
number of elements in the input.

@@

For surface-surface and surface-void functions we use an edge detection filter
and then calculate either autocorrelation of the edge or cross-correlation of
the edge and the void phase. The edge is extracted by convolving an input with a
short signal $E$ (\cref{eq:filter-1d} for 1D case, \cref{eq:filter-2d} for 2D case or
\cref{eq:filter-3d-1}-\cref{eq:filter-3d-3} for 3D case).
\begin{align}
  E^{1D} &= \frac{\sqrt{2}}{3} \left[
    \begin{array}{ccc}
        1 & -2 & 1
      \end{array}
    \right] \label{eq:filter-1d} \\
  E^{2D} &= \frac{\sqrt{2}}{9} \left[
    \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & -8 & 1 \\
        1 & 1 & 1
      \end{array}
    \right] \label{eq:filter-2d} \\
  E^{3D}_{ij0} &= \frac{\sqrt{2}}{27} \left[
    \begin{array}{ccc}
        1 & 1 & 1 \\
        1 & 1 & 1 \\
        1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3d-1} \\
  E^{3D}_{ij1} &= \frac{\sqrt{2}}{27} \left[
    \begin{array}{ccc}
      1 & 1 & 1 \\
      1 & -26 & 1 \\
      1 & 1 & 1
    \end{array}
    \right] \label{eq:filter-3d-2} \\
  E^{3D}_{ij2} &= E^{3D}_{ij0} \label{eq:filter-3d-3}
\end{align}

An algorithm for $F_{ss}$ function is the following:
\begin{algorithmic}[1]
  \Procedure{$F_{ss}$}{$A, phase, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $E_A \gets E * A^{(phase)}$
  \State $\hat{E}_A \gets F[E_A]$
  \State $\hat{F}_{ss} \gets |\hat{E}_A| / N$
  \State \textbf{return} $F^{-1} [\hat{F}_{ss}]$
  \EndProcedure
\end{algorithmic}

Using \cref{eq:cross-ft} we obtain an algorithm for surface-void function:
\begin{algorithmic}[1]
  \Procedure{$F_{sv}$}{$A, phase, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $A^{(void)} \gets I^{(void)} (A)$
  \State $E_A \gets E * A^{(phase)}$
  \State $\hat{E}_A \gets F[E_A]$
  \State $\hat{A}^{(void)} \gets F[A^{(void)}]$
  \State $\hat{F}_{sv} \gets \hat{E}_A \overline{\hat{A}^{(void)}} / N$
  \State \textbf{return} $F^{-1} [\hat{F}_{sv}]$
  \EndProcedure
\end{algorithmic}

\textcolor{red}{Lineal path function?}

>>>>

The next function is a cluster function $C_2$.
\begin{definition}
  Cluster function $C_2^{(i)}(\bm{r})$ equals to probability that both
  ends of a vector $\bm{r}$ lie in the same cluster when the vector is
  randomly thrown into the sample. A cluster is a set of points of the same
  phase $i$ where any two points can be connected by a path lying entirely in
  that phase.
\end{definition}

The next two functions are called surface-surface ($F_{ss}$) and surface-void
($F_{sv}$) functions and describe surface of a sample. We skip strict verbal
definitions and introduce an interface indicator function instead:
\begin{equation*}
  M^{(i)}(\bm{r}) = | \nabla I^{(i)}(\bm{r}) |
\end{equation*}
Here differentiation is understood in the sense of generalized functions,
because an ordinary differential of $I$ is either zero or undefined. For
homogenous media definitions of these functions are:
\begin{align}
  F_{ss}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) M^{(i)}(\bm{x} +
  \bm{r}) \rangle \label{eq:fss-def} \\
  F_{sv}^{(i)}(\bm{r}) &= \langle M^{(i)}(\bm{x}) I^{(void)}(\bm{x}
  + \bm{r}) \rangle \label{eq:fsv-def}
\end{align}
These functions are defined for media with dimensionality of 2 and
higher. $F_{sv}$ is a physical quantity inversely proportional to length and
$F_{ss}$ is inversionaly proportional to surface area, therefore we can use
$\mu m^{-1}$ to measure surface-void and $\mu m^{-2}$ to measure surface-surface
functions. Putting it simply, surface-void function measures length (2D case) or
surface (3D case) of a part of the interface which lies in the void phase of
porous medium shifted by vector $\bm{r}$. Surface-surface function characterizes
intersections of interfaces of porous medium and its shifted version. As always,
there are versions of these functions applicable to scalar argument for
isotropic media.

@@

Finally, a chord length function $p^{(i)}(r)$ is defined as follows:
\begin{definition}
$p^{(i)}(r)dz$ = Probability of finding a chord of length between $r$ and $r+dr$
in phase $i$. Chord is a line segment which lies entirely in the same phase and
touches the interface with its ends.
\end{definition}

\verb+CorrelationFunctions.jl+ is capable to compute all these
functions. Usually there are two implementations for a function. The first
implementation computes a correlation function in specific predefined directions
(i.e. orientation of test vectors thrown into a sample is fixed). There are one
predefined direction in 1D case, four directions in 2D case and thirteen
directions in 3D case. This implementation is somewhat slow but requires much
less memory than the second one. The second implementation computes a
correlation function in all possible directions and is faster but requires much
memory. These implementations are contained in \code{Directional} and \code{Map}
modules of the package correspondingly.

@@

\section{The map approach.}
\label{sec:map}
Definition \cref{eq:s2-def} can be rewritten as follows:
\begin{equation}
  S_2^{(i)}(r) = \frac{F^{-1} [|F [A^{(i)}](z)|] (r)}{N(r)} \label{eq:s2-ft}
\end{equation}
where $A^{(i)} \xleftarrow{I^{(i)}} A$ is a binary array obtained by element-wise
application of interface indicator $I^{(i)}$ to an input $A$, $F$ is a discrete
Fourier transform (DFT) operator and $N(r)$ is a total number of trials (vectors
thrown into the sample). If we want to calculate $S_2$ function in periodic mode
we need only to apply \cref{eq:s2-ft} to an input to obtain the result. In
non-periodic mode we need to pad an input with zeros to at least twice the size
minus one for each dimension. In periodic mode a total number of trials is
independent of $r$ and equals to the number of elements in an input array. In
non-periodic mode $N(r)$ is calculated as described \textcolor{red}{Here}. The
following algorithm summarizes all said above:
\begin{algorithmic}[1]
  \Procedure{$S_2$}{$A, phase, periodic$}
  \If{$\neg periodic$}
  \State $A \gets zeropad(A)$
  \Comment Pad with at least $2\cdot size(A, i) - 1$ zeros for each dimension $i$.
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $\hat{A}^{(phase)} \gets F[A^{(phase)}]$
  \State $\hat{S}_2 \gets |\hat{A}^{(phase)}| / N$
  \Comment $N$ is an array with numbers of trials
  \State \textbf{return} $F^{-1} [\hat{S}_2]$
  \EndProcedure
\end{algorithmic}
Algorithmic complexity of this implementation if $O(M \log M)$ where $M$ is the
number of elements in the input.

@@

For surface-surface and surface-void functions we use an edge detection filter
and then calculate either autocorrelation of the edge or cross-correlation of
the edge and the void phase. The edge is extracted by convolving an input with a
short low-pass filter $H$. It has a width $7$ and all its coefficients with
exception of the central coefficient are inversely proportional to distance
from the center:
\begin{equation}
  \begin{aligned}
    H_{ij} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2}} & \quad i \ne 3, j \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right. \\
    H_{ijk} &= S \left\{
    \begin{array}{cc}
      \frac{1}{\sqrt{(i-3)^2 + (j-3)^2 + (k-3)^3}} & \quad i \ne 3, j \ne 3, k
      \ne 3 \\
      C & \quad \text{otherwise}
    \end{array}
    \right.
  \end{aligned}
  \label{eq:filter-7x7}
\end{equation}
where $C$ is such that all coefficients of $H'$ sum to zero, and $S$ equals to
$30.45849$ in 2D and $172.96232$ in 3D case.

An algorithm for $F_{ss}$ function is the following:
\begin{algorithmic}[1]
  \Procedure{$F_{ss}$}{$A, phase, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $A_H \gets H * A^{(phase)}$
  \State $\hat{A}_H \gets F[A_H]$
  \State $\hat{F}_{ss} \gets |\hat{A}_H| / N$
  \State \textbf{return} $F^{-1} [\hat{F}_{ss}]$
  \EndProcedure
\end{algorithmic}

Using \cref{eq:cross-ft} we obtain an algorithm for surface-void function:
\begin{algorithmic}[1]
  \Procedure{$F_{sv}$}{$A, phase, periodic$}
  \If{$\neg periodic$}
    \State $A \gets zeropad(A)$
  \EndIf
  \State $A^{(phase)} \gets I^{(phase)} (A)$
  \State $A^{(void)} \gets I^{(void)} (A)$
  \State $A_H \gets H * A^{(phase)}$
  \State $\hat{A}_H \gets F[A_H]$
  \State $\hat{A}^{(void)} \gets F[A^{(void)}]$
  \State $\hat{F}_{sv} \gets \hat{A}_H \overline{\hat{A}^{(void)}} / N$
  \State \textbf{return} $F^{-1} [\hat{F}_{sv}]$
  \EndProcedure
\end{algorithmic}

Our library does not have an implementation for lineal-path maps, although we
know that such implementations exist.
